A decade of biodiversity change in an estuarine assemblage: fits and starts and natural events

### 2022-06-17 sub-projects/trawl/docs/ms/ms.RMD
Big commit. Updated qD models and plots. work on manuscript to finalize results

---
title: "A decade of biodiversity change in an estuarine assemblage: fits and starts and natural events"
author:
   - James R Junker:
     email: james.junker1@gmail.com
     institute: [glrc, lumcon]
     correspondence: yes
   - Craig R McClain:
     institute: lumcon
institute:
   - glrc: Great Lakes Research Center, Michigan Technological University, Houghton, MI USA
   - lumcon: Louisiana Universities Marine Consortium, Chauvin, LA USA
output:
   word_document:
      reference_docx: working_docx_template.docx
      toc: no
      pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
   pdf_document:
     keep_tex: true
   html_document: default
   pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
header-includes:
- \usepackage{lineno}
- \usepackage{amsmath}
- \usepackage{indentfirst}
- \linenumbers
indent: true
linestretch: 1
bibliography: refs.bib
link-citations: no
link-color: grey
csl: ecology.csl
---

```{r Rmarkdown setup, include = FALSE, echo=F, warning=FALSE, message=FALSE}
here::i_am("sub-projects/trawl/docs/ms/ms.RMD")

source(here::here("datascript.R"))
source(here::here("sub-projects/trawl/R/Lorenz.R"))
source(here::here("sub-projects/trawl/R/Evenness.R"))
# which years should we remove
year_removals = c("2007","2020","2021")

@@

TB_trawl_taxasite <- TB_trawl_data %>%
  dplyr::filter(year(Date) %ni% year_removals) %>%
  dplyr::mutate(Date = as.Date(paste0(as.character(year),"-",as.character(month),"-01"))) %>%
  dplyr::select(Date, date_id, Common_name, Abundance) %>%
  group_by(Date, date_id, Common_name) %>%
  summarise(Abundance = sum(Abundance)) %>%
  na.omit %>% group_by(Date, date_id, Common_name) %>%
  # # remove trawls with less than 10 individuals 
  # dplyr::filter(sum(Abundance) >=10) %>%
  # group_by(Date, date_id, Common_name) %>%
  pivot_wider(names_from = Common_name, values_from = Abundance, values_fill = list(Abundance = 0)) %>%
  ungroup() %>%
  dplyr::select(Date, date_id, everything())

@@

# # create bootstrapped data set to estimate variability
# TB_trawl_boots = TB_trawl_data %>%
#   group_by(year) %>%
#   modelr::bootstrap(TB_trawl_data)

# set specific variables used in methods
## how many bootstrap permutations to run with replacement? Keep low until final draft
nperm = 99
## how many individuals for count for Sn, Sn = 18 * n_year
n_mth = 2
#

@@

<!---  --->
Anthropogenic activities continue to exert long-term, varied, and increasing pressures on ecosystems [@steffen2015] causing alterations to biological diversity on local and global scales. Shifts in biodiversity at local scales are variable and potentially more nuanced than at global scales, but generally local changes manifest primarily as changes to community composition and structure rather than declines in richness [@hillebrand2018]. These shifts in communities through time are difficult to predict, in part, due to the multidimensional nature of biodiversity [@chase2018]. Yet, understanding how communities change through time is crucial for predicting current and future changes to biological systems and ecosystem services.
<!--- --->
Explaining the extent and pace of biodiversity change will require identifying the conditions that lead to divergent biological responses to environmental change, both natural and anthropogenic. In particular, it will be important to disentangle the roles of intrinsic (e.g., ecological) and extrinsic (e.g., environmental) drivers of biodiversity change. Intrinsic and extrinsic processes can interact, leading to temporal patterns of biodiversity change that vary in response to environmental change. For example, gradual environmental change over time can lead to gradual community shifts or these shifts may be undetectable or abrupt [@williams2011][@scheffer2003].

Disentangling the complexity of drivers and responses in community structure is especially pressing for marine systems. Marine systems have experienced greater changes than other habitats [@antao2020]. Yet, we currently lack linkages between ecological theory, primarily derived from terrestrial systems, and data in marine systems leaving unanswered questions about biodiversity changes in these habitats [@blowes2019]. We propose a framework to explore the nature of biodiversity change in an estuary bay community and connect this change to intrinsic and extrinsic drivers. This research will build towards a more coherent theory of biodiversity change through time [@shoemaker2020].

Here, we quantified biodiversity changes within a shallow estuarine community over a decade to understand shifts in total species richness, the relative distribution of abundant and rare species, and potential change points in community composition through time...

@@

This study was focused on Terrebonne Bay, located in the northern Gulf of Mexico of Louisiana, USA (29^$\circ$^11.200'N, 90^$\circ$^36.560'W; Figure S1). Terrebonne Bay is a broad (1,761 km2), shallow estuary (average depth 2m), separated from the Gulf of Mexico by two barrier island chains and connected by four tidal passes. The bay-estuary complex is part of an abandoned distributary of the Mississippi River and therefore no longer receives any significant riverine inputs. Thus, the majority of its freshwater inputs are from the local drainage basin [@bianchi2009]. While water column salinity and density are homogenous, strong salinity gradients exist seaward from the marsh to the barrier islands [@bianchi2009]. Tides in Terrebonne Bay are diurnal with a semidiurnal component, ranging 20-80 cm [@inoue2000].

@@

Trawl sampling was conducted as part of the Louisiana Universities Marine Consortium's (LUMCON) educational outreach program (https://www.lumcon.edu/education-and-outreach). Individual trawls were conducted using *trawl apparatus description...* for 10 - 20 min. To minimize the effects of trawl-to-trawl variability, we combined all trawls monthly.

@@

Species names were checked for typos and confirmed using the *gni_parse* function from the *taxize* package [@chamberlain2020] and further standardized to consistent taxonomic resolution for entire data series (Supplemental materials). To track trends in species richness through time, we calculated  effort-controlled  species richness, *S~n~*, which provides the expected number of species given a defined number of *n* randomly sampled individuals [@gotelli2001]. For annual and monthly *S~n~* calculations, *n* was set as the minimum number of individuals sampled. We also calculated the uncorrected accumulation of species over time by filtering species from the data set after first occurrence. We then calculated asymptotic species richness, *S~asymp~*, using the Chao1 estimator [@chao1984; @chao1987]. This metric is highly correlated with measured richness, *S* [@mcgill2011], but attempts to extrapolate the total species richness based on incomplete sampling of the community. We calculated *S~asymp~* and *S~n~* using the mobr package [@mcglinn2019; @mcglinn2021a].

@@

To test for changes in species richness over time, we fit bayesian implementations of generalized additive models (GAMs) using the `mgcv` [@wood2011] and  `brms` package in R [@burkner2017; @burkner2018].

@@

```{r env data, echo=FALSE, warning=FALSE, message=FALSE}

@@

@@

<!--- Outline the general patterns annually and  --->
Over the sampling period from `r min(TB_trawl_sampling$year, na.rm = TRUE)` to `r max(TB_trawl_sampling$year, na.rm = TRUE)`, `r length(unique(TB_trawl_data$Date))` trawls-days were performed yielding `r format(signif(sum(TB_trawl_data$Abundance, na.rm = TRUE),3), scientific = TRUE)` individuals. Among years, total effort generally increased, both in number of trawl-days and individuals counted, from a minimum effort of `r min(TB_trawl_sampling$trawl_days)` (`r TB_trawl_sampling %>% slice_min(trawl_days) %>% dplyr::select(year) %>% unlist`) to `r max(TB_trawl_sampling$trawl_days)` (`r TB_trawl_sampling %>% slice_max(trawl_days) %>% dplyr::select(year) %>% unlist`) days and from `r min(TB_trawl_sampling$individuals, na.rm = TRUE)` (`r TB_trawl_sampling %>% slice_min(individuals) %>% dplyr::select(year) %>% unlist`) to `r format(signif(max(TB_trawl_sampling$individuals, na.rm = TRUE),3), scientific = TRUE)` (`r TB_trawl_sampling %>% slice_max(individuals) %>% dplyr::select(year) %>% unlist`) individuals. In total, `r length(unique(TB_trawl_data$Common_name))` unique species were identified and new species were consistently observed throughout the sampling period (Figure S1). Within a calendar year, asymptotic species richness (*S~asymp~*) was relatively stable from 2009 to 2012, reaching a minimum value of `r yr_group_stats %>% dplyr::filter(index == "S_asymp") %>% slice_min(median) %>% dplyr::select(median) %>% unlist %>% as.numeric %>% round(0)` in `r yr_group_stats %>% dplyr::filter(index == "S_asymp") %>% slice_min(median) %>% dplyr::select(year)`. Estimated *S~asymp~* increased in 2013 to a series maximum of `r yr_group_stats %>% dplyr::filter(index == "S_asymp") %>% slice_max(median) %>% dplyr::select(median) %>% unlist %>% as.numeric %>% round(0)` in `r yr_group_stats %>% data.frame %>% dplyr::filter(index == "S_asymp") %>% slice_max(median) %>% dplyr::select(year)`. However, given the large differences in effort among years, variability in *S~asymp~* may be partly attributable to sampling effects among years. Therefore, we also calculated rarefied richness at `r S_n_yr` individuals, *S~`r S_n_yr`~*,  for each year. Here, effort-controlled species richness, *S~`r S_n_yr`~*, diverged qualitatively from *S~asymp~* and was generally constant with an apparent slight increase in more recent sampling years. Overall, *S~`r S_n_yr`~* ranged from `r yr_group_stats %>% dplyr::filter(index == "S_n_yr") %>% slice_min(median) %>% dplyr::select(median) %>% round(1)` to `r yr_group_stats %>% dplyr::filter(index == "S_n_yr") %>% slice_max(median) %>% dplyr::select(median) %>% round(1)` among all years. Monthly, of rarefied richness averaged `r round(mean(mth_sn$S_n, na.rm = TRUE),1)` [range: `r round(min(mth_sn$S_n, na.rm = TRUE),1)`--`r round(max(mth_sn$S_n, na.rm = TRUE),1)`] and exhibited seasonal oscillations, but no detectable trend over time (Figure 1a, Table S1).

```{r figure 1, echo = FALSE, warning = FALSE, message=FALSE, fig.cap="Figure 1. Multiple measures of species richness within the trawl community between the years 2007 to 2019 show continual accumulation of species (purple points and line) through the timeseries. The estimated asymptotic species richness over a calendar year (Sasymp; median values and 95% percentile interval represented by yellow points and area), suggests increasing numbers of species within the community over time. However, rarefied species richness controlled to 900 individuals (S900; median values and 95% percentil interval represented by green line and area) suggests the perceived increase in Sasymp may represent an effect of increased sampling effort"}

### create timeseries figure
mth_group_stats %>%
  dplyr::filter(index == 'S_n') %>%
  ggplot(aes( x= Date, y = median))+
  scale_y_continuous(name = expression("Rarefied richness ("*italic(S[n])*")"), limits = c(0,12))+
  scale_x_date(date_breaks = "1 year", date_labels = "%Y",limits = c(as.Date("2009-01-01"),as.Date("2019-12-31"))) +
  geom_point(size = 1.5) +
  geom_line(size = 1.2) 
  theme()

@@

```{r turnover, echo = FALSE, warning=FALSE, message=FALSE}

## quantify loss and appearance of species in the timeseries
trawl_mth_turnover = purrr::map(c('appearance','disappearance','total'),
                                     ~codyn::turnover(df = TB_trawl_taxasite_long,
                                                      time.var = 'date_id',
                                                      species.var = 'Common_name',
                                                      abundance.var = 'Abundance',
                                                      metric = .x) %>%
                                       data.frame %>%
                                       dplyr::mutate(Date = as.Date(paste0(unique(TB_trawl_yr$year)[-1],"-01-01"), format = "%Y-%m-%d"),
                                                     variable = .x) %>%
                                       dplyr::rename(value = !!names(.[1])) %>%
                                       bind_rows(data.frame(value = NA_real_, time.var = NA_real_, Date = as.Date("2008-01-01"), variable =.x))) %>%
                                     bind_rows
gains = trawl_yr_turnover %>%
  dplyr::filter(variable == "appearance")

losses = trawl_yr_turnover %>%
  dplyr::filter(variable == "disappearance")

total = trawl_yr_turnover %>%
  dplyr::filter(variable == "total")

@@

Consistent with the pattern of continued accumulation of new species throughout the sampling series and relatively stable effort-controlled richness, species gains (`r round(mean(gains$value, na.rm = TRUE),2)` $\pm$ `r round(sd(gains$value, na.rm = TRUE),2)` SD) and losses (`r round(mean(losses$value, na.rm = TRUE),2)` $\pm$ `r round(sd(losses$value, na.rm = TRUE),2)` SD) generally balanced over time (Figure 1b). Total turnover was variable but relatively stable (`r round(mean(total$value, na.rm = TRUE),2)` $\pm$ `r round(sd(total$value, na.rm = TRUE),2)` SD) and exhibited no detectable trend or coherent seasonal component.

```{r turnover plot ,echo = FALSE, warning=FALSE, message=FALSE, fig.cap = "Figure 2. Total species turnover and partitioning of species gains and losses over time."}
trawl_mth_turnover %>%
  na.omit %>%
  ggplot()+
  geom_line(aes(x = Date, y = value, group = variable, color = variable), size = 1.1)+
  scale_color_manual(values = viridis(n = 3, option = 'viridis'))+
  scale_y_continuous(name = 'Species turnover',limits = c(0,1), expand = c(0.01,0.01))+
  scale_x_date(date_breaks = 'year', limits = c(as.Date("2009-01-01"),as.Date("2019-12-31")),
               date_labels = "%Y")+
  theme(legend.position = c(0,1),
        legend.justification = c(0,1),
        legend.background = element_rect(fill = 'transparent', color = NA),
        legend.title = element_blank(),
        axis.title.x = element_blank())
```

```{r annual qD, echo=FALSE, warning=FALSE, message=FALSE}
set.seed(123)
q2_boots = boot_sample_groups(TB_trawl_taxasite %>% dplyr::select(-julian_date,-Date,-month,-biweekly), 
                   group_var = 'year',
                   qD = 2,
                   n_perm = nperm)
set.seed(123)
q1_boots = boot_sample_groups(TB_trawl_taxasite %>% dplyr::select(-julian_date,-Date,-month,-biweekly), 
                              group_var = 'year',
                              qD = 1,
                              n_perm = nperm)

>>>>

---
title: "A decade of biodiversity change in an estuarine assemblage: fits and starts and natural events"
author:
   - name: James R Junker
     email: james.junker1@gmail.com
     institute: [glrc, lumcon]
     correspondence: true
   - name: Craig R McClain
     institute: lumcon
institute:
   - glrc: Great Lakes Research Center, Michigan Technological University, Houghton, MI USA
   - lumcon: Louisiana Universities Marine Consortium, Chauvin, LA USA
output:
   word_document:
      reference_docx: working_docx_template.docx
      toc: no
      pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
   pdf_document:
     keep_tex: true
   html_document: default
   pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
header-includes:
- \usepackage{lineno}
- \usepackage{amsmath}
- \usepackage{indentfirst}
- \linenumbers
indent: true
linestretch: 1
bibliography: refs.bib
link-citations: no
link-color: grey
csl: ecology.csl
---

```{r Rmarkdown setup, include = FALSE, echo=F, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
here::i_am("sub-projects/trawl/docs/ms/ms.RMD")

source(here::here("datascript.R"))
source(here::here("sub-projects/trawl/R/Lorenz.R"))
source(here::here("sub-projects/trawl/R/Evenness.R"))

library(mgcv)
library(brms)
library(bayesplot)

rerunEven  = FALSE
rerunD = FALSE
# which years should we remove
year_removals = c("2007","2020","2021","2022")

@@

TB_trawl_taxasite <- TB_trawl_data %>%
  dplyr::mutate(Date = as.Date(paste0(as.character(year),"-",as.character(month),"-01"))) %>%
  dplyr::select(Date, date_id, Common_name, Abundance) %>%
  group_by(Date, date_id, Common_name) %>%
  dplyr::summarise(Abundance = sum(Abundance)) %>%
  na.omit %>% group_by(Date, date_id, Common_name) %>%
  # # remove trawls with less than 10 individuals 
  # dplyr::filter(sum(Abundance) >=10) %>%
  # group_by(Date, date_id, Common_name) %>%
  pivot_wider(names_from = Common_name, values_from = Abundance, values_fill = list(Abundance = 0)) %>%
  ungroup() %>%
  dplyr::select(Date, date_id, everything())

@@

TB_trawl_effort_mth = TB_trawl_data %>%
  dplyr::filter(year(Date) %ni% year_removals) %>%
  dplyr::select(Date, date_id) %>%
  group_by(date_id) %>%
  dplyr::summarise(trawl_n = n()) %>%
  dplyr::mutate(Date = as.Date(date_id, origin = "2007-01-01")) %>%
  dplyr::select(Date, trawl_n)

# set specific variables used in methods
## how many bootstrap permutations to run with replacement? Keep low until final draft
nperm = 99
## how many individuals for count for Sn, Sn = 18 * n_year
n_mth = 2
#
iter = 7000
warmup = 5000
thin = 10
adapt_delta = 0.99

@@

<!---  --->
Anthropogenic activities continue to exert long-term, varied, and increasing pressures on ecosystems [@steffen2015] causing alterations to biological diversity on local and global scales. Shifts in biodiversity at local scales are variable and potentially more nuanced than at global scales, but generally local changes manifest primarily as changes to community composition and structure rather than declines in richness [@hillebrand2018]. These shifts in communities through time are difficult to predict, in part, due to the multidimensional nature of biodiversity [@chase2018]. Yet, understanding how communities change through time is crucial for predicting current and future changes to biological systems and ecosystem services.

<!--- --->
Explaining the extent and pace of biodiversity change will require identifying the conditions that lead to divergent biological responses to environmental change, both natural and anthropogenic. In particular, it will be important to disentangle the roles of intrinsic (e.g., ecological) and extrinsic (e.g., environmental) drivers of biodiversity change. Intrinsic and extrinsic processes can interact, leading to temporal patterns of biodiversity change that vary in response to environmental change. For example, gradual environmental change over time can lead to gradual community shifts or these shifts may be undetectable or abrupt [@williams2011; @scheffer2003].

<!--- --->
Disentangling the complexity of drivers and responses in community structure is especially pressing for marine systems. Marine systems have experienced greater changes than other habitats [@antao2020]. Yet, we currently lack linkages between ecological theory, primarily derived from terrestrial systems, and data in marine systems leaving unanswered questions about biodiversity changes in these habitats [@blowes2019]. We propose a framework to explore the nature of biodiversity change in an estuary bay community and connect this change to intrinsic and extrinsic drivers. This research will build towards a more coherent theory of biodiversity change through time [@shoemaker2020].

<!--- --->
Here, we quantified biodiversity changes within a shallow estuarine community over a decade to understand shifts in total species richness, the relative distribution of abundant and rare species, and potential change points in community composition through time...

@@

This study was focused on Terrebonne Bay, located in the northern Gulf of Mexico of Louisiana, USA (29^$\circ$^11.200'N, 90^$\circ$^36.560'W; Figure S1). Terrebonne Bay is a broad (1,761 km2), shallow estuary (average depth ~2m), separated from the Gulf of Mexico by two barrier island chains and connected by four tidal passes. The bay-estuary complex is part of an abandoned distributary of the Mississippi River and therefore no longer receives significant riverine inputs. Thus, the majority of its freshwater inputs are from the local drainage basin [@bianchi2009]. While water column salinity and density are homogeneous, strong salinity gradients exist seaward from the marsh to the barrier islands [@bianchi2009]. Tides in Terrebonne Bay are diurnal with a semidiurnal component, ranging 20-80 cm [@inoue2000].

@@

Trawl sampling was conducted as part of the Louisiana Universities Marine Consortium's (LUMCON) educational outreach program (https://www.lumcon.edu/education-and-outreach) from 2009 to 2019. Individual trawls were conducted using *trawl apparatus description...* for 10 - 20 min. To dampen potentially volatile trawl-to-trawl variability, we combined all trawls monthly and therefore report changes based on monthly aggregated trawl sampling.

@@

Species names were checked for typos and confirmed using the *gni_parse* function from the *taxize* package [@chamberlain2020] and further standardized to consistent taxonomic resolution for the entire data series (Supplemental materials). To track trends in species richness through time, we calculated  effort-controlled  species richness, *S~n~*, which provides the expected number of species given a defined number of *n* randomly sampled individuals [@gotelli2001]. For annual and monthly *S~n~* calculations, *n* was set as the minimum number of individuals sampled. We also calculated the uncorrected accumulation of species over time by filtering species from the data set after first occurrence. We then calculated asymptotic species richness, *S~asymp~*, using the Chao1 estimator [@chao1984; @chao1987]. This metric is highly correlated with measured richness, *S* [@mcgill2011], but attempts to extrapolate the total species richness based on incomplete sampling of the community. We calculated *S~asymp~* and *S~n~* using the mobr package [@mcglinn2019; @mcglinn2021a].

@@

To detect trends in species richness and other dimensions of biodiversity (e.g., turnover, evenness, etc.) we used generalized additive models (GAMs) with a bayesian implementations using the `mgcv` [@wood2011] and  `brms` packages in R [@burkner2017; @burkner2018]. Full models included alinear term for trawl date and smoothing spline terms for month:
$$y = \beta_0 + \beta_1 * time + f_{season}(x_1) + \varepsilon,\;\;\; \varepsilon ~ N(0, \sigma^{2} \Lambda) $$
where, $y$ is the response variable (e.g., richness, turnover, evenness, etc.), $\beta_0$ is an intercept, $\beta_1$ a coefficient for trawl date, and $f_{season}$ represents seasonal smooth functions for month of year. Seasonal terms were fit with cyclic cubic regression splines with smoothing dimension *k* = 12 to allow for continuity between December and January among model years.

@@

@@

mth_even = TB_trawl_taxasite %>%
  dplyr::select(-Date, -date_id) %>%
  apply(., 1, gini_even) %>%
  t() %>%
  data.frame() %>%
  bind_cols(TB_trawl_taxasite[,c('Date','date_id')], .) %>%
  setNames(., c("Date","date_id", "gini_raw", "gini_norm")) %>%
  dplyr::mutate(month = month(Date)) %>%
  # pivot_longer(-Date:-date_id, names_to = "measure", values_to = "median") %>%
  ungroup

# mth_even %>%
#   ggplot(aes(x = Date, y = gini_norm))+
#   geom_point()+
#   geom_line()+
#   scale_y_continuous(name = 'Normalized Gini coefficient', limits = c(0,0.5))+
#   scale_x_date(date_breaks = 'year', limits = c(as.Date("2009-01-01"),as.Date("2019-12-31")),
#                date_labels = "%Y")

mth_group_stats = mth_group_stats %>%
  bind_rows(mth_even %>%
              pivot_longer(-Date:-date_id, names_to = 'index', values_to = 'median'))

## quantify loss and appearance of species in the timeseries
trawl_mth_turnover = purrr::map(c('appearance','disappearance','total'),
                                ~codyn::turnover(df = TB_trawl_taxasite_long %>% dplyr::select(-Date),
                                                 time.var = 'date_id',
                                                 species.var = 'Common_name',
                                                 abundance.var = 'Abundance',
                                                 metric = .x) %>%
                                  data.frame %>%
                                  dplyr::mutate(date_id = date_id,
                                                variable = .x) %>%
                                  dplyr::rename(value = !!names(.[1])) %>%
                                  bind_rows(data.frame(value = NA_real_, variable =.x))) %>%
  bind_rows %>%
  left_join(TB_trawl_taxasite[,c(1,2)],.,by = 'date_id')

gains = trawl_mth_turnover %>%
  dplyr::filter(variable == "appearance")

losses = trawl_mth_turnover %>%
  dplyr::filter(variable == "disappearance")

total = trawl_mth_turnover %>%
  dplyr::filter(variable == "total")

## add trendline in for A and B

@@

```{r evenness, echo = FALSE, warning=FALSE, message=FALSE }
if(rerunEven){
source(here::here("sub-projects/trawl/R/run_evenness_models.R"))
} else if(!all(file.exists(c(here::here("sub-projects/trawl/derived-data/even_null.rds"),
                             here::here("sub-projects/trawl/derived-data/even_s.rds"),
                             here::here("sub-projects/trawl/derived-data/even_full.rds"))))){
  source(here::here("sub-projects/trawl/R/run_evenness_models.R"))
} else{
  even_bayes_full_mod = readRDS(here::here("sub-projects/trawl/derived-data/even_full.rds"))
  even_bayes_s_mod = readRDS(here::here("sub-projects/trawl/derived-data/even_s.rds"))
  even_bayes_null_mod = readRDS(here::here("sub-projects/trawl/derived-data/even_null.rds"))
}

@@

<!--- Outline the general patterns annually and  --->
Over the sampling period from `r min(TB_trawl_sampling_yr$year, na.rm = TRUE)` to `r max(TB_trawl_sampling_yr$year, na.rm = TRUE)`, `r length(unique(TB_trawl_data$Date))` trawl-days were performed yielding `r format(signif(sum(TB_trawl_data$Abundance, na.rm = TRUE),3), scientific = TRUE)` individuals. Among years, total effort generally increased, both in number of trawl-days and individuals counted, from a minimum effort of `r min(TB_trawl_sampling_yr$trawl_days)` (`r TB_trawl_sampling_yr %>% slice_min(trawl_days) %>% dplyr::select(year) %>% unlist`) to `r max(TB_trawl_sampling_yr$trawl_days)` (`r TB_trawl_sampling_yr %>% slice_max(trawl_days) %>% dplyr::select(year) %>% unlist`) days and from `r min(TB_trawl_sampling_yr$individuals, na.rm = TRUE)` (`r TB_trawl_sampling_yr %>% slice_min(individuals) %>% dplyr::select(year) %>% unlist`) to `r format(signif(max(TB_trawl_sampling_yr$individuals, na.rm = TRUE),3), scientific = TRUE)` (`r TB_trawl_sampling_yr %>% slice_max(individuals) %>% dplyr::select(year) %>% unlist`) individuals. In total, `r length(unique(TB_trawl_data$Common_name))` unique species were identified and new species were consistently observed throughout the sampling period (Figure S1). Within a calendar year, asymptotic species richness (*S~asymp~*) was relatively stable from 2009 to 2012, reaching a minimum value of `r yr_group_stats %>% dplyr::filter(index == "S_asymp") %>% slice_min(median) %>% dplyr::select(median) %>% unlist %>% as.numeric %>% round(0)` in `r yr_group_stats %>% dplyr::filter(index == "S_asymp") %>% slice_min(median) %>% dplyr::select(year)`. Estimated *S~asymp~* increased in 2013 to a series maximum of `r yr_group_stats %>% dplyr::filter(index == "S_asymp") %>% slice_max(median) %>% dplyr::select(median) %>% unlist %>% as.numeric %>% round(0)` in `r yr_group_stats %>% data.frame %>% dplyr::filter(index == "S_asymp") %>% slice_max(median) %>% dplyr::select(year)`. However, given the large differences in effort among years, variability in *S~asymp~* may be partly attributable to sampling effects among years. Therefore, we also calculated rarefied richness at `r S_n_yr` individuals, *S~`r S_n_yr`~*,  for each year. Here, effort-controlled species richness, *S~`r S_n_yr`~*, diverged qualitatively from *S~asymp~* and was generally constant with an apparent slight increase in more recent sampling years. Overall, *S~`r S_n_yr`~* ranged from `r yr_group_stats %>% dplyr::filter(index == "S_n") %>% slice_min(median) %>% dplyr::select(median) %>% round(1)` to `r yr_group_stats %>% dplyr::filter(index == "S_n") %>% slice_max(median) %>% dplyr::select(median) %>% round(1)` among all years. Monthly, of rarefied richness averaged `r round(mean(mth_Sn$S_n, na.rm = TRUE),1)` [range: `r round(min(mth_Sn$S_n, na.rm = TRUE),1)`--`r round(max(mth_Sn$S_n, na.rm = TRUE),1)`] and exhibited seasonal oscillations, but no detectable trend over time (Figure 1A, Table S1).

```{r figure 1, echo = FALSE, warning = FALSE, message=FALSE, fig.height=8.5, fig.width=8.5, fig.cap="Figure 1. A) Rarefied species richness within the nekton community between the years 2009 to 2019 show variable, but consistent, numbers of species over time. B) Despite no trend in the total number of species, the trawl community showed continual turnover throughout."}

## Goals estimate the rarified richness through time
## Show changes in evenness and Hill #s

### create timeseries figure
mth_group_stats %>%
  dplyr::filter(index == 'S_n') %>%
  ggplot(aes( x= Date, y = median))+
  scale_y_continuous(name = expression("Rarefied richness ("~italic(~S[n])~")"), limits = c(1,12),
                     breaks = c(1,5,10),expand = c(0.01,0.01))+
  scale_x_date(date_breaks = "year", limits = c(as.Date("2009-01-01"),as.Date("2019-12-31")),
               date_labels = "%Y",expand = c(0.01,0.01))+
  geom_point(size = 2)+
  geom_line(size = 1.2)+
  theme_tufte()+
  geom_rangeframe(sides = "l")+
  annotate('text', label = "A", x = as.Date("2009-01-30"), y = 11, family ='serif', size = 8, hjust = 0)+
  theme(axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        # panel.grid.minor.x = element_line(color = 'lightgray'),
        plot.margin = unit(c(0.0,0.1,-0.1,0.0), 'cm'), 
        plot.background = element_rect(fill = 'transparent', color = NA),
        panel.background = element_rect(fill = 'transparent', color = NA),
        axis.title.y = element_text(family = 'serif', size = 16),
        axis.text = element_text(family = 'serif', size = 14)) -> rich_time;

trawl_mth_turnover %>%
  dplyr::filter(variable == 'total') %>%
  na.omit %>%
  ggplot(aes(x = Date, y = value, group = variable, color = variable))+
  geom_point(size = 2)+
  geom_line(size = 1.2)+
  scale_color_manual(values = viridis(n = 1, option = 'viridis', begin = .75))+
  scale_y_continuous(name = 'Species turnover',limits = c(0,1), expand = c(0.01,0.01),
                     breaks = c(0,0.2 ,0.4, 0.6, 0.8), labels = c(0,0.2,0.4,0.6,0.8))+
  scale_x_date(date_breaks = 'year', limits = c(as.Date("2009-01-01"),as.Date("2019-12-31")),
               date_labels = "%Y",expand = c(0.01,0.01))+
  theme_tufte()+
  geom_rangeframe(sides = "lb", color = 'black')+
  annotate('text', label = "B", x = as.Date("2009-01-30"), y = 1, family = 'serif', vjust = 1, size = 8, hjust = 0)+
  theme(legend.position = "none",
        legend.justification = c(0,1),
        axis.title.x = element_blank(),
        plot.margin = unit(c(-0.18,0.15,0.0,0.0), 'cm'),
        # panel.grid.minor.x = element_line(color = 'lightgray'),
        plot.background = element_rect(fill = 'transparent', color = NA), 
        axis.title.y = element_text(family = 'serif', size = 16),
        axis.text = element_text(family = 'serif', size = 14)) -> turnover_time

g2 <- ggplotGrob(rich_time)
g3 <- ggplotGrob(turnover_time)
g2$widths <- unit.pmax(g2$widths, g3$widths)
g3$widths <- unit.pmax(g2$widths, g3$widths)
grid.arrange(g2,g3, layout_matrix = rbind(1,1,1,1,2,2))

@@

```{r diversity2 plot, echo = FALSE, warning=FALSE, message=FALSE, fig.cap= "Figure 2. A) Temporal trends in the effective number of species of select orders of diversity (q = 0.1, 1,2). Effective species are presented as percentages of ^0^D, or species richness. ^0.1^D highlights the trends in the percentage of species considered very rare, while ^1^D treats rare and dominant species equally, and ^2^D highlights the number of dominant species in the community through time. ^1^D is equivalent to exp(Shannon-Weiner H) and ^2^D is equivalent to 1/Simpson's diversity index. B) The estimated temporal trend for ^q^D orders (q = 0.1--3) from conditional effects of time in generalized additive models. The black line represents the mean estimate surrouned by 25^th^ and 75^th^ (dark grey) and 5^th^ and 95^th^ percentiles (light grey) of the posterior distribution."}

#### patterns of Hill diversity through time}
## hill

hill_time = function(x,...){
  hillR::hill_taxa(TB_trawl_taxasite %>% dplyr::select(-Date,-date_id), q = x) %>%
    data.frame %>% dplyr::mutate(q = paste0("q",as.character(x))) %>% setNames(.,c('value','q')) %>%
    bind_cols(TB_trawl_taxasite %>% dplyr::select(Date, date_id),.)
}
# debugonce(hill_time)
hill_time_df = furrr::future_map(seq(0,3, by = 0.1), ~hill_time(.x)) %>%
  bind_rows %>%
  group_by(Date, date_id) %>% 
  pivot_wider(names_from = 'q', values_from = 'value') %>%
  dplyr::mutate(across(contains('q'), list(~(.x/q0)*100), .names = "{.col}_eff")) %>%
  dplyr::mutate(month = month(Date)) %>%
  dplyr::select(date_id, month, contains('eff'))

eff_cols = names(hill_time_df)[grepl("*eff", names(hill_time_df))]

model_qD = function(x,...){
  form_string = paste(x," ~ date_id + s(month, bs = 'cc', k = 12)")
  bayes_form = bf(formula = formula(form_string))
  mod = brm(bayes_form,
            data = hill_time_df,
            family = gaussian(),
            cores = 4, seed = 42,
            iter = 30000, warmup = 25000, thin = 10,
            control = list(adapt_delta = 0.99),
            save_pars = save_pars(all = TRUE),
            file = paste0(here::here("sub-projects/trawl/derived-data"),"/",as.character(x),"_full.rds"),
            file_refit = 'on_change')

  return(posterior_summary(mod, pars = 'date_id', probs = c(0.025,0.25,0.5,0.75,0.975)))
}

if(rerunD){
  # debugonce(model_qD)
eff_cols %>%
    furrr::future_map(~model_qD(.x)) %>%
    setNames(., nm = eff_cols) %>%
    saveRDS(file = here::here("sub-projects/trawl/derived-data/qD_coefs.rds"))
}

qD_coefs = readRDS( file = here::here("sub-projects/trawl/derived-data/qD_coefs.rds"))
qD_coefs_df = qD_coefs %>% lapply(., data.frame) %>% bind_rows(.id = "model") %>%
  dplyr::mutate(q = as.numeric(sub("^q(\\d{1}|\\d{1}.\\d{1}).*","\\1", model)))

q0.1_bayes_full_mod = readRDS(file = here::here("sub-projects/trawl/derived-data/q0.1_eff_full.rds"))

ceff0.1 <- conditional_effects(q0.1_bayes_full_mod, 'date_id')

ceff0.1_df = data.frame(
  date_id = ceff0.1$date_id$date_id,
  qD = "q0.1_eff",
  estimate = ceff0.1$date_id$estimate__,
  upper = ceff0.1$date_id$upper__,
  lower = ceff0.1$date_id$lower__
)

q0.5_bayes_full_mod = readRDS(file = here::here("sub-projects/trawl/derived-data/q0.5_eff_full.rds"))

ceff0.5 <- conditional_effects(q0.5_bayes_full_mod, 'date_id')

ceff0.5_df = data.frame(
  date_id = ceff0.5$date_id$date_id,
  qD = "q0.5_eff",
  estimate = ceff0.5$date_id$estimate__,
  upper = ceff0.5$date_id$upper__,
  lower = ceff0.5$date_id$lower__
)

q2_bayes_full_mod = readRDS(file = here::here("sub-projects/trawl/derived-data/q2_eff_full.rds"))

ceff2 <- conditional_effects(q2_bayes_full_mod, 'date_id')
